---
title: "WHO Health Indicators - West Africa"
subtitle: "Statistical Analysis Report"
author: "Abdel Kader Younoussi Saley"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5
)
source("analysis.R")
```

# Overview

This report presents a statistical analysis of WHO health indicators for the
15 West African countries. Data is ingested automatically from the
WHO Global Health Observatory API via the Python pipeline and exported
to CSV for R-based analysis.

**Indicators covered:**

- Estimated malaria deaths
- Under-five mortality rate (per 1,000 live births)
- Maternal mortality ratio (per 100,000 live births)
- Nurses and midwives per 10,000 population
- Tuberculosis incidence (per 100,000 population)
- Wasting prevalence in children under 5 (%)

---

# Data Summary

```{r load-data}
df <- load_who_data()
```

```{r summary-table}
kable(
  df %>%
    group_by(indicator) %>%
    summarise(
      Countries    = n_distinct(country),
      Years        = paste(min(year), "-", max(year)),
      Observations = n(),
      Mean         = round(mean(value, na.rm = TRUE), 2),
      Min          = round(min(value, na.rm = TRUE), 2),
      Max          = round(max(value, na.rm = TRUE), 2)
    ),
  caption = "Descriptive statistics by indicator"
)
```

---

# Correlation Analysis

The correlation matrix below examines relationships between health indicators
at country level, using the most recent available observation per country.

```{r correlation-plot, fig.height=7}
cor_result <- compute_correlation(df)
plot_correlation(cor_result$cor_matrix)
```

---

# Trend Analysis - Niger

Niger is selected as a focus country given its representative profile
among low-income West African states. A linear trend model is fitted
for each available indicator.

## Under-five Mortality Rate

```{r niger-u5mr}
trend <- fit_country_trend(df, "Niger", "MDG_0000000007")
plot_trend(trend, "Niger", "Under-five mortality rate (per 1,000 live births)")
```

```{r niger-u5mr-stats}
if (!is.null(trend)) {
  kable(
    trend$tidy %>% select(term, estimate, std.error, statistic, p.value) %>%
      mutate(across(where(is.numeric), ~ round(., 4))),
    caption = "Linear model coefficients - Niger U5MR"
  )
}
```

## Maternal Mortality Ratio

```{r niger-mmr}
trend_mmr <- fit_country_trend(df, "Niger", "MDG_0000000026")
plot_trend(trend_mmr, "Niger", "Maternal mortality ratio (per 100,000 live births)")
```

---

# Regional Summary

```{r regional-table}
regional_tbl <- build_regional_table(df)
kable(
  regional_tbl %>% select(country, indicator, latest_year, latest_value, trend),
  caption = "Latest values and trend direction - West African countries"
)
```

---

# Methodology Notes

**Data source:** WHO Global Health Observatory OData API
(`https://ghoapi.azureedge.net/api`)

**Pipeline:** Data ingested via Python (`pipeline.py`), exported to CSV,
loaded here via `readr`. This separation ensures reproducibility:
the R analysis is independent of the Python environment.

**Trend model:** Ordinary least squares regression (`lm`) with year as
predictor. Slope direction (positive/negative) used to classify trends.
Statistical significance assessed at p < 0.05.

**Correlation:** Pearson correlation computed on latest available
observation per country per indicator, using pairwise complete observations.

---

*Report generated automatically. Data reflects WHO GHO as of last pipeline run.*
